Ages of Stars and Planets in the Kepler Field Over the First Four Billion Years

### 2024-08-26 papers/paper/ms.tex
update ms (caveat discn rework)

\subsubsection{A Paucity Of Young Stars And Planets}

The stellar ages are given in Table~\ref{tab:planets} for the known
planet hosts, and in Table~\ref{tab:stars} for all Kepler stars.
Figure~\ref{fig:hist_tgyro} shows the rotation-based age distributions
of all the stars (left column) and the KOI hosts (middle column).
These ``histograms'' are sampled from the posterior probability
distributions by drawing ten random samples from each posterior, and
then computing the normalized histogram of the resulting samples.
We also considered an alternative approach for constructing these
plots using hierarchical Bayesian deconvolution
\citep{Masuda2022infer}, and found similar results.  The plotted
uncertainties assume Poisson statistics.  We truncated the plots
at $4$\,Gyr, which is the upper bound for our
rotational age calibration.  The overall slope yields a paucity of
young stars relative to the expectation of a uniform age distribution.

@@

\added{{\it Temperature sensitivity}---There are quantitative differences
in Figure~\ref{fig:hist_tgyro} between the
K dwarfs (4400--5400\,K) and the FGK stars (3800--6200\,K).  For instance,
the distribution peaks at 2.3\,Gyr for K dwarfs, and at 2.9\,Gyr for
the FGK sample.  The peak is also sharper for the K dwarfs;
its decrease from 2.3-3\,Gyr is in an age range in which we believe the
rotation period catalog completeness to be near unity.  We
discuss further details of the $t_{\rm gyro}$--$T_{\rm eff}$
distributions in Appendix~\ref{app:age_diagnostic}, and propose possible
explanations for these differences in Section~\ref{subsec:caveats}. }

@@

Most of the stars observed by Kepler are in the thin disk.  This can
be verified by following \citet{Gaia_2018}, and labeling stars with 2D
tangential velocities $v_{\rm T}$$<$40\,\kms\ as thin disk members,
and those with 60$<$$v_{\rm T}$$<$150\,\kms\ as thick disk members.
For all stars in the KIC, this yields 95{,}694 thin and 54{,}039 thick
disk stars respectively.  The thin disk stars have a threefold larger
detected rotation period fraction: 35{,}675 rotators are in the thin
disk, while 7{,}312 are in the thick disk.  Further imposing the
gyrochronology quality flags discussed in Section~\ref{subsec:flags}
yields 17{,}755 thin-disk stars for which gyrochronology is
applicable, and 2{,}462 thick-disk stars.

@@

The right column of Figure~\ref{fig:hist_tgyro} compares our derived
age distribution against star formation histories (SFHs) reported by
\citet{2019A&A...624L...1M} and \citet{2020NatAs...4..965R} based on
CMD fitting.  The overall slope of both SFHs broadly agrees with what
we find from rotation-based ages.  The SFH from
\citet{2019A&A...624L...1M} seems entirely consistent, particularly
after accounting for the statistical uncertainties of that study.
Regarding the episodic star formation bursts reported by
\citet{2020NatAs...4..965R}, they are not apparent in our overall FGK
star sample (top row).  However, rotation-based ages are most precise
for $\approx$G8V-K4V dwarfs older than $\approx$1.5\,Gyr
\citep{Bouma_2023}.  The lower row of Figure~\ref{fig:hist_tgyro}
selects these stars with a temperature cut, and may show a hint of the
$\approx$2\,Gyr spike reported by \citet{2020NatAs...4..965R}.  The
$\approx$1\,Gyr spike however is not recovered.  This
$\approx$2.5\,Gyr local maximum is similarly present in the isochrone
ages derived by \citet{Berger_2020a_catalog} for the Kepler field, and
in the red giant asteroseismic ages derived by
\citet{SilvaAguirre2018}.  The main novelty of our ages relative to
those studies is their improved accuracy at $<$1\,Gyr.

@@

Our most constraining ages tend to come from one source of
information: rotation.  Factors other than \deleted{stellar }age and mass can
influence \deleted{stellar }rotation rates.  \added{Astrophysical factors include
metallicity, differential rotation, and tidal spin-up, while observational
challenges include the difficulty of measuring small-amplitude,
long-period signals that are only quasi-coherent due to the starspot
lifetime approaching the stellar rotation period.  In the following, we attempt to
quantify the importance of these effects in our analysis, ordered by
what we perceive to be their descending importance.}

\begin{figure*}[!t]
	\begin{center}
		\leavevmode
%		\subfloat{
%			\includegraphics[width=.98\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240821_maxage4000.pdf}
%		}
%		
%		\vspace{-0.35cm}
%		\subfloat{
%			\includegraphics[width=.98\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240821_maxage4000_preciseagesonly.pdf}
%		}
		%
		%	\vspace{-0.35cm}
			\subfloat{
				\includegraphics[width=.98\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240821_maxage4000_dropfracshortrot.pdf}
			}
	\end{center}
	\vspace{-0.6cm}
	\caption{{\bf Variant of Figure~\ref{fig:hist_tgyro}}, with rotation-derived age distributions that
			correct for the estimated (estm.) effect of unresolved $\lesssim$0.5\,AU binaries. 
		Even after applying our quality cuts (Section~\ref{subsec:flags}), up to
		\fivepctnuniqstarsantosrotgyroappl\ out of \ngyroappllttendays\ ``clean'' rotators with
		$P_{\rm rot}$$<$10\,days could be unresolved binaries.
		These distributions correct for this possibility by randomly removing
		what we expect to be the maximum possible number of such systems; the actual
		correction relative to Figure~\ref{fig:hist_tgyro} might be smaller.
		\label{fig:hist_tgyro_corr}
	}
\end{figure*}

{\it Binaries}:
\replaced{Even intermediate-separation ($\sim$10\,AU) binaries
are biased toward rapid rotation \citep[e.g.][and many studies
thereafter]{Meibom_2007}.}{Binaries are often biased
toward rapid rotation, even at intermediate separations of $\lesssim$10\,AU \citep[e.g.][]{Meibom_2007,Gruner_2023}.}
\added{\citet{2019ApJ...871..174S} for instance found that 59\% of
Kepler stars with rotation periods between 1.5 and 7\,days lie
0.3\,mag above the main sequence, compared with 28\% of their full
rotation sample.
We attempted to mitigate the effects of binarity with five distinct
quality flags (Section~\ref{subsec:flags}), and this attempt was at least
somewhat effective, based on the peak near zero in the raw
distributions of Figure~\ref{fig:hist_tgyro} being flattened in the
cleaned distributions.
However, the sensitivity of our quality flags is hard to assess.  One specific concern is that these
quality flags might have minimal sensitivity to binaries with
separations of 0.05-0.5\,AU, for all but the highest mass ratios.  This inner limit is set by the onset
of detectable ellipsoidal variations in \citet{2016AJ....151...68K}; the outer limit is set by the typical 
orbital periods of astrometric binaries in Gaia DR3 \citep{GaiaCollaboration2023}.  }

\added{ If we were not at all sensitive to
binaries with separations of 0.05-0.5\,AU, then based on
the binary fraction and separation distribution from
\citet{Raghavan2010}, as much as $\approx$5\% of our ``gyro
applicable'' sample
(\fivepctnuniqstarsantosrotgyroappl/\nuniqstarsantosrotgyroappl\
stars) might be contaminated.  In a worst case scenario, all of
these unidentified binaries might be able to induce $P_{\rm rot}$$<$10
days, which would then be mixed in with the bona fide young, single stars.
In our actual sample of apparently clean stars,
\ngyroappllttendays/\nuniqstarsantosrotgyroappl\ had $P_{\rm
rot}$$<$10\,days.

Figure~\ref{fig:hist_tgyro_corr} shows the rotation-based age distribution
calculated by randomly dropping \fivepctnuniqstarsantosrotgyroappl\ of the
$P_{\rm rot}$$<$10\,day stars.
There are three important aspects of this result, relative to the original
distributions in Figure~\ref{fig:hist_tgyro}.
The first is that it would imply a much larger drop in the star formation rate,
by a factor of 4.81$\pm$0.26, rather than \ratiosfr$\pm$\uncratiosfr,
over the past three gigayears.
The second is that it allows for a constant rate of decline in star formation
since $\approx$2.8\,Gyr ago, rather than having a ``kink'' at 1.5\,Gyr,
as in Figure~\ref{fig:hist_tgyro}.
And finally, these corrected distributions show closer agreement with SFHs
derived from CMD fitting.
We caution however that this correction assumes that {\it none} of our
quality flags were able to identify any of these unresolved binaries.
Since we did in fact include flags based on e.g.~location in the HR diagram,
this correction is best interpreted as an upper limit on the possible impact of binarity.}

%TODO TODO TODO
\added{{\it Erroneous period measurements:} }

\added{
{\it Metallicity:}
}

\added{
{\it Differential rotation:}
}

%TODO TODO TODO
%TODO TODO TODO
%TODO TODO TODO

Metallicity may also be relevant
\citep{2020MNRAS.499.3481A,2024arXiv240500779S}.  An additional caveat
is that our rotation-based ages used photometric effective
temperatures (Section~\ref{sec:stellarprops}), even though
spectroscopic temperatures are available for the planet-hosting
subset.  This decision was driven by a desire for homogeneity 
irrespective of planet-hosting status.  However it implies that the
estimated planet ages might shift if one were to account for the
spectroscopic information.

We were interested in not only the overall age distribution of the
Kepler field, but also in the reliability of individual ages for young
stars known to host planets.  We attempted to assess this reliability using
quality bitmasks (Table~\ref{tab:planets}), a comparison against \deleted{open }
cluster ages (Section~\ref{sec:litagecomparison}), and a comparison
between rotation and lithium ages.  The \deleted{open }cluster comparison
suggested that our ages were generally accurate at $\ll$0.5\,dex
across 0.04--2.5\,Gyr, similar to their quoted precision.
Nonetheless, the upper-left panel of Figure~\ref{fig:agescalecompone}
does show outliers, typically $\lesssim$10\% of the population
in any cluster.  These stars could be either field star contaminants,
or else anomalous stars whose physical rotation histories were altered
by processes not captured by our statistical uncertainties.
Similarly, although we found $t_{\rm gyro}$ and $t_{\rm Li}$ to agree
for \ltonegyrhighqconfirmedtwosided\ planets orbiting ``high-quality''
stars, we did find one system, Kepler-786, with radically different
lithium (\kepseveneightsix\,Myr) and rotational
($\approx$4.4\,Gyr) ages.  Planetary mergers
are one process that could produce such a signal, but
testing this would require a method for measuring differential abundances
across a large number of elements.

\added{A separate plausible origin for rotational outliers is that they come from
erroneously reported rotation periods.  While
the large surveys do show good internal consistency (Appendix~\ref{app:mcqonly}),
the referee correctly pointed out that this is only a comparative test of methods,
rather than an ability to recover ground truth \citep[e.g.][]{Aigrain2015}.
One could imagine an adversarial scenario in which, say, 10-20\%\ of stars
with $P_{\rm rot}$$>$20\,days are erroneously reported, and in fact have much
longer rotation periods.  In this scenario, our analysis would overestimate the
degree by which the SFR in the Kepler field has decreased, by 7-14\%. }

Other age indicators may help in verifying the ages of the
$\lesssim$3\,Gyr stars that are the \added{main }focus of this work.  Specifically,
chromospheric emission in the X-ray, \ion{Ca}{2} HK line, and the UV
can serve as an age tracer
\citep{Mamajek_2008,2014MNRAS.441.2361V,2024ApJ...960...62E}.
However, these age indicators are all in a sense ``rotation-powered''.
The dynamo converts kinetic energy into magnetic energy, which is
emitted through these chromospheric pathways.  We did not attempt to
incorporate these indicators into this study due to concerns regarding
sensitivity, homogeneity, and the question of whether they in fact
provide age information that is truly independent from rotation.

@@

\added{{\it Ages of stars older than the Sun}: }\deleted{A broader question, beyond
the scope of this work, concerns the ages of stars $\gtrsim$4\,Gyr
old.}\added{This work was limited to stars younger than 4\,Gyr}. For
age-dating models based only on rotation rates, an important challenge
is that the detectability of the rotation signals for 3-10\,Gyr stars,
especially Sun-like stars, is at the limits of the Kepler data
\citep{2022ApJ...937...94M}.  In the ``old star'' regime, methods that
leverage either asteroseismology
\citep{vanSaders_2016,2024ApJ...962..138S}, or else that combine both
the evolution of stellar luminosity and stellar rotation
\citep{Angus_2019,Claytor2020,2023ApJ...952..131M} seem the most
capable of providing useful age constraints for single stars over the
full span of their main-sequence lifetimes.

@@

{\it Searches for new associations}:
Some of the youngest stars in Tables~\ref{tab:planets}
and~\ref{tab:stars} have been linked to their birth clusters, while
others have not. A systematic search for the birth clusters of the
youngest ``field stars'' could combine positions and velocities from
Gaia with rotation measurements from TESS. Given the $\sim$100\,Myr
decoherence time of young clusters, the youngest ``field stars'' may
have easily identifiable young neighbors.

@@

Tables~\ref{tab:planets} and~\ref{tab:stars} summarize the results for
planets and stars, respectively.  Our recovered ages are accurate for
all 14 known Kepler planets in clusters.  While lithium provided
minimal added information for most of the sample, $t_{\rm Li}$ and
$t_{\rm gyro}$ agreed in $\gtrsim$90\% of cases for which comparison
was possible.  Our resulting ages included two-sided $t_{\rm gyro}$
and $t_{\rm Li}$ for \ltonegyrhighqconfirmedtwosided\ sub-gigayear
confirmed planets, and two-sided $t_{\rm gyro}$ with one-sided $t_{\rm
Li}$ for \ltonegyrhighqconfirmedonesided\ sub-gigayear confirmed
planets.  Allowing for ``candidate'' planets, grazing transit
geometries, and stars with RUWE far from unity expands the counts by a
factor of two.

The sizes and periods for the most secure set of planets are shown in
Figure~\ref{fig:rp_period_age_results}.   While the new young planets
are mostly mini-Neptunes, some are near the lower boundary of the
``sub-Jovian desert'' \citep{Owen2018}, which could have an
evolutionary connection to their youth.  Other discoveries include
Earth-sized planets with new ages of only a few hundred million years.

@@

This analysis is particularly interesting in relation to the question
of how fast stars $\gtrsim$4\,Gyr old spin down.  If we lived in a
universe in which stellar rotation rates and spot-induced photometric
amplitudes remained {\it constant} after $\approx$3\,Gyr, then older
stars \replaced{would tend to }{could }contaminate the 2.3-3\,Gyr peaks in
Figures~\ref{fig:hist_tgyro} and~\ref{fig:hist_tgyro_mcq}.  This
scenario deserves consideration, given\replaced{ evidence from asteroseismic
rotation periods that has pointed to}{ {\it i)} asteroseismic evidence for} slowed spin-down rates after
$\approx$4\,Gyr for Sun-like stars
\citep[e.g.][]{vanSaders_2016,2024ApJ...962..138S}\added{, and {\it ii)} the observation of a ``long-period edge''
in the $P_{\rm rot}$-$T_{\rm eff}$ distribution which
may pile up over similar timescales \citep{David_2022}}.  However, \replaced{because
the \citeauthor{McQuillan_2014} sample does not include many stars
older than $\approx$4\,Gyr \citep{2022ApJ...937...94M}, this
analysis}{this pileup is thought to happen at $>$5\,Gyr for
late-G dwarfs and early K dwarfs \citep[see Figure~4
by][]{2024ApJ...962..138S}.  This means that it cannot affect the
2.3\,Gyr peak in the age distribution that we see for 4400-5400\,K stars.   
Moreover, while some contamination by older stars may be present in the peaks,
the analysis by \citet{2022ApJ...937...94M} fitted
isochrones to a high-quality subset of Kepler data, and found that the
\citeauthor{McQuillan_2014} sample does not
include many stars older than $\approx$\,4\,Gyr.}
\replaced{would argue}{These lines of evidence suggest} that the peak
\replaced{cannot be}{in the age distribution is not} caused by a pile-up of
much older stars with especially lethargic spin-down rates\added{, but that it could be
contaminated by them, especially for stars hotter than the Sun}.

>>>>

\subsubsection{Kepler's Demographic Cliff}

The stellar ages are given in Table~\ref{tab:planets} for the known
planet hosts, and in Table~\ref{tab:stars} for all Kepler stars.
Figure~\ref{fig:hist_tgyro} shows the rotation-based age distributions
of all the stars (left column) and the KOI hosts (middle column).
These ``histograms'' are sampled from the posterior probability
distributions by drawing ten random samples from each posterior, and
then computing the normalized histogram of the resulting samples.
We also considered an alternative approach for constructing these
plots using hierarchical Bayesian deconvolution
\citep{Masuda2022infer}, and found similar results.  The plotted
uncertainties \replaced{assume Poisson statistics}{are Poissonian}.  We truncated the plots
at $4$\,Gyr, which is the upper bound for our
rotational age calibration.  The overall slope \replaced{yields}{shows} a paucity of
young stars relative to the expectation of a uniform age distribution.

@@

\added{{\it Temperature sensitivity}---There are quantitative differences
in Figure~\ref{fig:hist_tgyro} between the
K dwarfs (4400--5400\,K) and the FGK stars (3800--6200\,K).  For instance,
the distribution peaks at 2.3\,Gyr for K dwarfs, and at 2.9\,Gyr for
the FGK sample.  The peak is also sharper for the K dwarfs;
its decrease from 2.3-3\,Gyr is in an age range in which we believe the
rotation period catalog completeness to be near unity.  We
show further details of the $t_{\rm gyro}$--$T_{\rm eff}$
distributions in Appendix~\ref{app:age_diagnostic}, and propose possible
explanations for these differences in Section~\ref{subsec:caveats}. }

@@

Most \deleted{of the }stars observed by Kepler are in the thin disk.  This can
be verified by following \citet{Gaia_2018}, and labeling stars with 2D
tangential velocities $v_{\rm T}$$<$40\,\kms\ as thin disk members,
and those with 60$<$$v_{\rm T}$$<$150\,\kms\ as thick disk members.
\replaced{For all stars in the KIC, this yields }{These criteria yield }95{,}694 thin and 54{,}039 thick
disk stars \replaced{respectively}{observed by Kepler}.  The thin disk stars have a threefold larger
detected rotation period fraction: 35{,}675 rotators are in the thin
disk, while 7{,}312 are in the thick disk.  Further imposing the
\deleted{gyrochronology }quality flags discussed in Section~\ref{subsec:flags}
yields 17{,}755 thin disk stars for which gyrochronology is
applicable, and 2{,}462 thick disk stars.

@@

The right column of Figure~\ref{fig:hist_tgyro} compares our derived
age distribution against star formation histories (SFHs) reported by
\citet{2019A&A...624L...1M} and \citet{2020NatAs...4..965R} based on
CMD fitting.  The overall slope of both SFHs broadly agrees with what
we find from rotation-based ages.  The SFH from
\citet{2019A&A...624L...1M} seems entirely consistent, particularly
after accounting for the statistical uncertainties of that study.
Regarding the episodic star formation bursts reported by
\citet{2020NatAs...4..965R}, they are not apparent in our overall FGK
star sample (top row).  However, rotation-based ages are most precise
for $\approx$G8V-K4V dwarfs older than $\approx$1.5\,Gyr
\citep{Bouma_2023}.  The lower row of Figure~\ref{fig:hist_tgyro}
selects these stars with a temperature cut, and may show a hint of the
$\approx$2\,Gyr spike reported by \citet{2020NatAs...4..965R}.  The
$\approx$1\,Gyr spike however is not recovered.  This
$\approx$2.5\,Gyr local maximum is similarly present in the isochrone
ages derived by \citet{Berger_2020a_catalog} for the Kepler field, and
in the red giant asteroseismic ages derived by
\citet{SilvaAguirre2018}.  The main novelty of our ages relative to
\replaced{those studies is their improved accuracy}{previous work is that they are demonstrably accurate} at $<$1\,Gyr\added{ (see Figure~\ref{fig:agescalecompone})}.

@@

\begin{figure*}[!t]
	\begin{center}
		\leavevmode
		\subfloat{
			\includegraphics[width=.98\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240821_maxage4000_dropfracshortrot.pdf}
		}
	\end{center}
	\vspace{-0.6cm}
	\caption{{\bf Potential impact of unresolved binaries.}   This figure shows
		the same raw (thin blue line) and quality-flag cleaned (``gyro applicable'') distributions
		as in Figure~\ref{fig:hist_tgyro}.
		However, even after applying our quality cuts (Section~\ref{subsec:flags}), up to
		\fivepctnuniqstarsantosrotgyroappl\ out of \ngyroappllttendays\ ``clean'' rotators with
		$P_{\rm rot}$$<$10\,days could be unresolved $\lesssim$0.5\,AU  binaries.
		The thick blue ``5\% binary contamination'' distribution corrects for this possibility by randomly removing
		what we estimate (estm.) to be the maximum possible number of such systems; the true
		correction might be smaller.  We expect such a correction to be more important in
		the sample of all Kepler stars than in the KOI host sample (see text).
		\label{fig:hist_tgyro_corr}
	}
\end{figure*}

Our most constraining ages tend to come from one source of
information: rotation.  Factors other than \deleted{stellar }age and mass can
influence \deleted{stellar }rotation rates.  \added{Astrophysical factors include
metallicity, differential rotation, mergers, and tidal spin-up.  Observational
limitations include the difficulty of measuring small-amplitude,
long-period signals that are only quasi-coherent due to the starspot
lifetime approaching the stellar rotation period.  There is also the issue that
our age-dating approach is not calibrated beyond 4\,Gyr.
In the following, we attempt to
quantify the importance of these effects in our analysis, ordered by
what we perceive to be their descending importance.}

{\it Binaries}:
\replaced{Even intermediate-separation ($\sim$10\,AU) binaries
are biased toward rapid rotation \citep[e.g.][and many studies
thereafter]{Meibom_2007}.}{Stars in binaries are often biased
toward rapid rotation, even beyond the typical
realm of tidal synchronization \citep[see][]{Meibom_2007,Gruner_2023}.}
\added{\citet{2019ApJ...871..174S} for instance found that 59\% of
Kepler stars with rotation periods between 1.5 and 7\,days lie
0.3\,mag above the main sequence, compared with 28\% of their full
rotation sample.
We attempted to mitigate the effects of binarity with five distinct
quality flags (Section~\ref{subsec:flags}).   This attempt was at least
somewhat effective, based on the peak near zero in the raw
distributions of Figure~\ref{fig:hist_tgyro} being flattened in the
cleaned distributions.
However, the completeness of our quality flags is hard to assess.  One specific concern is that these
flags might have minimal sensitivity to binaries with
separations of $\approx$0.05-0.5\,AU, for all but the highest mass ratios.  This inner limit is set by the onset
of detectable ellipsoidal variations in \citet{2016AJ....151...68K}; the outer limit is set by the typical 
orbital periods of astrometric binaries in Gaia DR3 \citep[see Figure~3 of][]{GaiaCollaboration2023}.  }

\added{ If we were not at all sensitive to
binaries with separations of 0.05-0.5\,AU, then based on
the binary fraction and separation distribution from
\citet{Raghavan2010}, as much as $\approx$5\% of our ``gyro
applicable'' sample
(\fivepctnuniqstarsantosrotgyroappl/\nuniqstarsantosrotgyroappl\
stars) might be contaminated.  In a worst case scenario, all of
these unidentified binaries might induce $P_{\rm rot}$$<$10
days, which would then be mixed in with the bona fide young, single stars.
This would constitute a major source of contamination,
given that our nominal sample of apparently clean stars had
\ngyroappllttendays\ stars with $P_{\rm
rot}$$<$10\,days.

Figure~\ref{fig:hist_tgyro_corr} shows the rotation-based age distribution
calculated by randomly dropping \fivepctnuniqstarsantosrotgyroappl\ of the
$P_{\rm rot}$$<$10\,day stars.
Relative to the nominal ``gyro applicable'' age distribution, there are three important differences.
There is a much larger drop in the star formation rate over the past three gigayears,
by a factor of 4.81$\pm$0.26, rather than \ratiosfr$\pm$\uncratiosfr.
The rate of decline in star formation
since $\approx$2.8\,Gyr ago is also constant, rather than having a ``kink'' at 1.5\,Gyr.
And finally, the corrected distributions show closer agreement with SFHs
derived from CMD fitting.
We caution however that this correction assumes that {\it none} of our
quality flags were able to identify any of these unresolved binaries.
Since we did in fact include flags based on e.g.,~location in the HR diagram,
this correction is best interpreted as an upper limit on the possible impact of binarity.}

\added{{\it Rotational outliers:} }
We were interested in not only the overall age distribution of the
Kepler field, but also in the reliability of individual ages for \replaced{young
stars known to host planets}{each star in our sample}.  We attempted to assess this reliability using
quality bitmasks (Table~\ref{tab:planets}), a comparison against \deleted{open }cluster ages (Section~\ref{sec:litagecomparison}), and a comparison
between rotation and lithium ages.  The \deleted{open }cluster comparison
suggested that our ages were generally accurate at $\ll$0.5\,dex
across 0.04--2.5\,Gyr, similar to their quoted precision.
Nonetheless, the upper-left panel of Figure~\ref{fig:agescalecompone}
does show outliers, typically $\lesssim$10\% of the population
in any cluster.  \added{While most of these outliers are known to have bad quality flags, some do not.} 
These stars could be either field star contaminants,
or else anomalous stars whose physical rotation histories were altered
by processes not captured by our statistical uncertainties.
Similarly, although we found consistent $t_{\rm gyro}$ and $t_{\rm Li}$
for \ltonegyrhighqconfirmedtwosided\ planets orbiting ``high-quality''
stars, we did find one system, Kepler-786, with radically different
lithium (\kepseveneightsix\,Myr) and rotational
($\approx$4.4\,Gyr) ages.  Planetary mergers
are one process that could produce such a signal, but
testing this would require a method for measuring differential abundances
across a large number of elements.

\added{A separate plausible origin for rotational outliers is that they come from
	erroneously reported rotation periods.  While
	the large surveys do show good internal consistency (Appendix~\ref{app:mcqonly}),
	the referee correctly pointed out that this is only a comparative test of methods,
	rather than an ability to recover ground truth \citep[e.g.][]{Aigrain2015}.
	One could imagine an adversarial scenario in which, say, 10-20\%\ of stars
	with $P_{\rm rot}$$>$20\,days are erroneously reported, and in fact have much
	longer rotation periods.  In this scenario, a similar analysis to that in Figure~\ref{fig:hist_tgyro_corr} showed
	that our analysis would overestimate the
	degree by which the SFR in the Kepler field has decreased, by 7-14\%. }

\added{Other age indicators may help in verifying the ages of the
$\lesssim$3\,Gyr stars that are the main focus of this work.  Specifically,
chromospheric emission in the X-ray, \ion{Ca}{2} HK line, and the UV
can serve as an age tracer
\citep{Mamajek_2008,2014MNRAS.441.2361V,2024ApJ...960...62E}.
However, these age indicators are all in a sense ``rotation-powered''.
The dynamo converts kinetic energy into magnetic energy, which is
emitted through these chromospheric pathways.  We did not attempt to
incorporate these indicators into this study due to concerns regarding
sensitivity, homogeneity, and the question of whether they in fact
provide age information that is truly independent from rotation.}

\added{{\it Weakened magnetic braking:} 
If we lived in a
universe in which stellar rotation rates and spot-induced photometric
amplitudes remained {\it constant} after $\approx$3\,Gyr, then older
stars could contaminate the peaks in
Figure~\ref{fig:hist_tgyro}.  This
scenario deserves consideration, given asteroseismic evidence for slowed spin-down rates after
$\approx$4\,Gyr for Sun-like stars
\citep[][]{vanSaders_2016,Hall2021,2024ApJ...962..138S}, and the observation of a ``long-period edge''
	in the $P_{\rm rot}$-$T_{\rm eff}$ distribution which
	may pile up over similar timescales \citep{David_2022}.  However, 
	this pileup is thought to happen at $>$5\,Gyr for
	late-G dwarfs and early K dwarfs \citep[see Figure~4
	by][]{2024ApJ...962..138S}.  This means that it cannot explain the
	2.3\,Gyr peak that we see in the 4400-5400\,K age distribution in Figure~\ref{fig:hist_tgyro}. 
	Independently,
	the analysis by \citet{2022ApJ...937...94M} fitted
	isochrones to a high-quality subset of Kepler data, and found that the
	\citet{McQuillan_2014} sample, discussed in Appendix~\ref{app:mcqonly}, does not
	include many stars older than $\approx$\,4\,Gyr.
These lines of evidence suggest that the peak
in the observed age distribution is not caused by a pile-up of
much older stars with especially lethargic spin-down rates, but that it could be
	contaminated by them beyond $\approx$3\,Gyr, especially for stars hotter than the Sun.
}

\added{{\it Metallicity:}  
Stars with higher metallicity are thought to have longer convective turnover
times, which may slow their rate of spin-down
	\citep{Amard2020}.
While observational studies of this effect have been suggestive, further work
remains to definitively establish the importance of the effect \citep[e.g.][]{Amard2020b,2024arXiv240500779S}
Nonetheless, based on the theoretical predictions, we flagged stars with known
	spectroscopic $|[{\rm Fe/H}]|$$>$0.3, based on either LAMOST or Keck/HIRES
	spectra \citep{Zong2018,Petigura_2022}.
The main limitation of this approach is that only 53.3\% of the stars in Table~\ref{tab:stars} with
$t_{\rm gyro}$$<$4\,Gyr had such spectra available.
For the rest of the sample, primarily composed of fainter stars,
this means that our quality flags were not sensitive to metallicity outliers.
Based on the metallicity distribution of stars for which we do have spectra,
this 
implies that at most $\approx$5\% of stars in our sample might have $|[{\rm Fe/H}]|$$>$0.3,
with a slight bias toward lower metallicity stars over higher metallicity.
Based on the results by \citet{Claytor2020},  this will be most important
for stars near the Kraft break, which could have ages over-estimated
by $\approx$50-100\% if they are in fact metal-poor, and under-estimated in the converse case.
}

\added{{\it Differential rotation:}
The Sun's surface rotates once per $\approx$25-30\,days at latitudes between the equator
and $\approx$$\pm$60$^\circ$ \citep{Snodgrass1990}.
This latitudinal surface differential rotation yields two separate limitations on our rotational ages.
The first is
on the accuracy with which any one star's rotation period can be measured over a few-year mission like Kepler;
the second is on the degree to which the intrinsic scatter in open cluster rotation sequences can be measured \citep[e.g.][]{Epstein_2014}.
In this study, we accounted for the latter intrinsic scatter based on available measurements
from Ruprecht-147 \citep{Curtis_2020} and M67 data
\citep{Barnes2016,Dungee_2022,Gruner_2023}.
This means that irrespective of the assumed $P_{\rm rot}$ measurement precision,
our model imposed a $\approx$5-10\% level of intrinsic $P_{\rm rot}$ scatter (perhaps differential-rotation induced)
at old ages.
While this accounts for the statistical impact of differential rotation on the inferred
age distribution,
the accuracy issue remains;
any one individual star's observed rotation period might be faster
or slower than average stars of the same mass and age.  
This effect can yield a $\approx$20\% age bias for individual Sun-like stars with ages
of a few billion years; the statistical uncertainties in Table~\ref{tab:stars} should be
treated with appropriate caution.
}

\deleted{Metallicity may also be relevant
\citep{2020MNRAS.499.3481A,2024arXiv240500779S}.  An additional caveat
is that our rotation-based ages used photometric effective
temperatures (Section~\ref{sec:stellarprops}), even though
spectroscopic temperatures are available for the planet-hosting
subset.  This decision was driven by a desire for homogeneity 
irrespective of planet-hosting status.  However it implies that the
estimated planet ages might shift if one were to account for the
spectroscopic information.}

\deleted{Other age indicators may help in verifying the ages of the
$\lesssim$3\,Gyr stars that are the \added{main }focus of this work.  Specifically,
chromospheric emission in the X-ray, \ion{Ca}{2} HK line, and the UV
can serve as an age tracer
\citep{Mamajek_2008,2014MNRAS.441.2361V,2024ApJ...960...62E}.
However, these age indicators are all in a sense ``rotation-powered''.
The dynamo converts kinetic energy into magnetic energy, which is
emitted through these chromospheric pathways.  We did not attempt to
incorporate these indicators into this study due to concerns regarding
sensitivity, homogeneity, and the question of whether they in fact
provide age information that is truly independent from rotation.}

@@

\added{{\it Ages of stars older than the Sun}: }\deleted{A broader question, beyond
the scope of this work, concerns the ages of stars $\gtrsim$4\,Gyr
old.}\added{This work was limited to stars younger than 4\,Gyr}. For
age-dating methods based only on rotation rates, an important challenge
is that the detectability of the rotation signals for 3-10\,Gyr stars,
especially Sun-like stars, is at the limits of the Kepler data
\citep{2022ApJ...937...94M}.  In the ``old star'' regime, methods that
leverage either asteroseismology
\citep{vanSaders_2016,2024ApJ...962..138S}, or else that combine both
the evolution of stellar luminosity and stellar rotation
\citep{Angus_2019,Claytor2020,2023ApJ...952..131M} seem the most
capable of providing useful age constraints for single stars over the
full span of their main-sequence lifetimes.

@@

{\it Searches for new associations}:
Some of the youngest stars in Tables~\ref{tab:planets}
and~\ref{tab:stars} have been linked to their birth clusters.
Others have not. A systematic search for the birth clusters of the
youngest ``field stars'' could combine positions and velocities from
Gaia with rotation measurements from TESS. Given the $\sim$100\,Myr
decoherence time of young clusters, the youngest ``field stars'' may
have easily identifiable young neighbors.

@@

Tables~\ref{tab:planets} and~\ref{tab:stars} summarize the results for
planets and stars, respectively.  Our recovered ages are accurate for
all 14 known Kepler planets in clusters.  While lithium provided
minimal added information for most of the sample\added{ due to its astrophysical
rate of depletion}, $t_{\rm Li}$ and
$t_{\rm gyro}$ agreed in $\gtrsim$90\% of cases for which comparison
was possible.  Our \replaced{resulting ages}{results} included two-sided $t_{\rm gyro}$
and $t_{\rm Li}$ for \ltonegyrhighqconfirmedtwosided\ sub-gigayear
confirmed planets, and two-sided $t_{\rm gyro}$ with one-sided $t_{\rm
Li}$ for \ltonegyrhighqconfirmedonesided\ sub-gigayear confirmed
planets.  Allowing for ``candidate'' planets, grazing transit
geometries, and stars with RUWE far from unity expands the counts by a
factor of two.

The sizes and periods for the most secure set of planets are shown in
Figure~\ref{fig:rp_period_age_results}.   While the new young planets
are mostly mini-Neptunes, some are near the lower boundary of the
``sub-Jovian desert'' \citep{Owen2018}, which could have an
evolutionary connection to their youth.  Other discoveries include
Earth-sized planets \added{such as Kepler-1312~c and Kepler-1561~b }with new ages of only a few hundred million years.

@@
